---
title: DeepGS combining R and Python
authors:
  - name: Markus Owen
format: 
  html:
    embed-resources: true
    code-fold: true
    number-sections: true
    toc: true
    toc-depth: 3
    date: now
    date-modified: last-modified
    date-format: "MMMM DD, YYYY, HH:mm:ss"
  pdf: 
    number-sections: true
    colorlinks: true
execute:
  echo: true
engine: knitr
# jupyter: python3
#engine: jupyter
#jupyter: ir
---

# Introduction

This Quarto document relates to [A deep convolutional neural network approach for predicting phenotypes from genotypes](https://doi.org/10.1007/s00425-018-2976-9). 

To get R and Python to work together I installed 

`install.packages("reticulate")` to allow Python to work from R, and 
`install.packages("IRkernel")` and `IRkernel::installspec()` to allow R to work from Jupyter. 

# R

```{r}
library(reticulate)
use_condaenv("gsc_tc_model", required = TRUE)
```

Which specific R is being executed? 

```{r}
file.path(R.home("bin"), "R")
```

## [rrBLUP mixed.solve example](https://cran.r-project.org/web/packages/rrBLUP/refman/rrBLUP.html#mixed.solve)

Create a random population of 200 lines with 1000 markers and random marker effects; use rrBLUP's `mixed.solve` to fit the mixed model of marker effects (using `Z=M`) and breeding values (using `K=A.mat(M)`). 

```{r}
library(rrBLUP)

m <- 1000
n <- 200
#random population of 200 lines with 1000 markers
M <- matrix(rep(0,n*m),n,m)
for (i in 1:n) {
  M[i,] <- ifelse(runif(m)<0.5,-1,1)
}
rownames(M) <- 1:n
A <- A.mat(M)

#random phenotypes
u <- rnorm(m)
g <- as.vector(crossprod(t(M),u))
h2 <- 0.5  #heritability
y <- g + rnorm(n,mean=0,sd=sqrt((1-h2)/h2*var(g)))

data <- data.frame(y=y,gid=1:n)

#predict marker effects
ans1 <- mixed.solve(y,Z=M)  #By default K = I
accuracy <- cor(u,ans1$u)

#predict breeding values
ans2 <- mixed.solve(y,K=A.mat(M))
accuracy <- cor(g,ans2$u)
```


Use `kin.blip` to predict breeding values, which should give identical values.  

```{r}
#predict breeding values using kin.blup, gives identical results to mixed.solve with K=A.mat(M)
ans3 <- kin.blup(data=data,geno="gid",pheno="y",K=A)
accuracy <- cor(g,ans3$g)
```


```{r}
# plot predictions based on marker effects
pred1 <- crossprod(t(M),ans1$u) + as.vector(ans1$beta)
plot(y,pred1)
abline(0,1,col='red')
```

```{r}
# plot predictions based on breeding value
pred2 <- ans2$u + as.vector(ans2$beta)
plot(y,pred2)
abline(0,1,col='red')
```

These two are identical, because the method to estimate the $u$ uses the kinship matrix anyway? 

```{r}
plot(pred1,pred2)
abline(0,1,col='red')
```


# python

Check that we can also fit the PyTorch CNN model. 

```{python}
# package imports
import torch
import torch.nn as nn
import torch.optim as optim
from torch.utils.data import TensorDataset, DataLoader

import os
import numpy as np
from datetime import datetime
import matplotlib.pyplot as plt
from sklearn import linear_model
import rdata

# local imports
from deepgs_model import DeepGSModel
from diagnostics import plot_actual_vs_predicted, evaluate_model, plot_both_vs_marker
import data_loader
from train_deepgs import train_deepGSModel,train_deepGSModel2,train_model
from config import cnnFrame

import run_training
```

```{python}
converted = rdata.read_rda("../data/wheat_example.rda")
```

```{python}
seed = 0
cvIdx = 0

# This call to load_wheat_data uses default rngSeed=0 so is reproducible
trainMat, trainPheno, validMat, validPheno, testMat, testPheno, trainIdx, validIdx, testIdx, markerImage = data_loader.load_wheat_data(cvIdx=cvIdx)
```

```{python}
device = "cuda" if torch.cuda.is_available() else "cpu"
if torch.backends.mps.is_available():
    device = torch.device("mps")
else:
    print ("MPS device not found.")

seed = 2
torch.manual_seed(seed)

model = train_deepGSModel(
    trainMat=trainMat,
    trainPheno=trainPheno,
    validMat=validMat,
    validPheno=validPheno,
    markerImage=markerImage,
    cnnFrame=cnnFrame,
    device=device,
    eval_metric="mae",
    num_round=10,
    batch_size=32,
    learning_rate=0.001,
    momentum=0.1,
    patience=10000
)

print("Training complete.")
print(model)
```


Make predictions with the fitted model: 

```{python}
model.eval()
# Train predictions
train_true, train_pred = evaluate_model(
    model, trainMat, trainPheno, markerImage, device=device)
# Train predictions
valid_true, valid_pred = evaluate_model(
    model, validMat, validPheno, markerImage, device=device)
# Train predictions
test_true, test_pred = evaluate_model(
    model, testMat, testPheno, markerImage, device=device)
model.train()
```

```{python}
plt.plot(train_true,train_pred,'o',c='g',label='Train')
plt.plot(valid_true,valid_pred,'^',c='b',label='Validation')
plt.plot(test_true,test_pred,'+',c='r',label='Test')
phenoMin = min(trainPheno.min(),validPheno.min(),testPheno.min())
phenoMax = max(trainPheno.max(),validPheno.max(),testPheno.max())
plt.plot([phenoMin, phenoMax],
            [phenoMin, phenoMax],
            'r--', lw=2)
plt.xlabel("Actual phenotype")
plt.ylabel("Predicted phenotype")
plt.title("Predicted vs Actual, cvIdx=%d" % cvIdx)
plt.grid(True)
plt.legend()
plt.show()
```

# Back to R - using rrBLUP with the wheat data

Extract the wheat data from Python variables using `py$`, and used rrBLUP's `mixed.solve` to fit marker effects. 

```{r}
py$converted

M = py$converted$wheat_example$M$values

y = py$converted$wheat_example$y$values

#predict marker effects
ans4 <- mixed.solve(y,Z=M)  #By default K = I

pred4 <- crossprod(t(M),ans4$u) + as.vector(ans4$beta)
```

## Plot the predictions based on marker effects vs the true values. 

```{r}
cor4 <- cor(y,pred4)
plot(y,pred4,main=paste("Wheat grain length, rrBLUP, Pearson correlation=", cor4))
abline(0,1,col='red')
```

## Plot the predictions based on breeding values vs the true values. 

```{r}
# predict breeding values (should be similar to phenotypes)
ans5 <- mixed.solve(y,K=A.mat(M))
cor5 <- cor(y,ans5$u)
plot(y,ans5$u,main=paste("Wheat grain length, rrBLUP with A.mat, Pearson correlation=", cor5))
abline(0,1,col='red')

```

# Using BGLR

```{r}
library('BGLR')
```

From "[Various Ways of fitting a 'GBLUP' model using BGLR](https://github.com/gdlc/BGLR-R/blob/master/inst/md/GBLUP.md#various-ways-of-fitting-a-gblup-model-using-bglr)"

## Fit the model 

```{r}
 data(wheat)

 nIter=12000
 burnIn=2000

 X=scale(wheat.X)/sqrt(ncol(wheat.X))
 y=wheat.Y[,1]

 fm1=BGLR( y=y,ETA=list(mrk=list(X=X,model='BRR')),
	   nIter=nIter,burnIn=burnIn,saveAt='brr_'
 	 )

 varE=scan('brr_varE.dat')
 varU=scan('brr_ETA_mrk_varB.dat')
 h2_1=varU/(varU+varE)
```

Check it works using plain matrix / numeric data. 

```{r}
 X=as.matrix(scale(wheat.X)/sqrt(ncol(wheat.X)))
 y=as.numeric(wheat.Y[,1])

 fm1=BGLR( y=y,ETA=list(mrk=list(X=X,model='BRR')),
	   nIter=nIter,burnIn=burnIn,saveAt='brr_'
 	 )

 varE=scan('brr_varE.dat')
 varU=scan('brr_ETA_mrk_varB.dat')
 h2_1=varU/(varU+varE)
```

Plot BGLR prediction vs true values. 

```{r}
pred6 <- predict(fm1)
cor6 <- cor(y,pred6)
plot(y,pred6,main=paste("Wheat grain length, BGLR, Pearson correlation=", cor6))
abline(0,1,col='red')
```

Predict marker effects and breeding values using rrBLUP instead

```{r}
#predict marker effect
ans7 <- mixed.solve(y,Z=wheat.X)
# predictions based on marker effects
pred7 <- crossprod(t(wheat.X),ans7$u) + as.vector(ans7$beta)
cor7 <- cor(y,pred7)


#predict breeding values
ans8 <- mixed.solve(y,K=A.mat(wheat.X))
# predictions based on breeding value
pred8 <- ans8$u + as.vector(ans8$beta)
cor8 <- cor(y,pred8)
plot(y,pred8,main=paste("Wheat grain length, rrBLUP, Pearson correlation=", cor7))
abline(0,1,col='red')

```

> Ridge regression best linear unbiased predictor and gBLUP were implemented using the package ‘rrBLUP’, with gBLUP using a marker-based relationship matrix (Endelman, 2011), whereas the Bayesian models were implemented using the package ‘BGLR’ (Pérez & de los Campos, 2014).

However, DeepGS implements GBLUP using BGLR!

### Plot GBLUP (BGLR) vs rrBLUP predictions

```{r}
plot(pred7,pred6,main="Wheat grain length",xlab="rrBLUP",ylab="GBLUP (Bayesian, BGLR)")
abline(0,1,col='red')
```

